The Magnetorotational Instability Prefers Three Dimensions

### 2019-07-23 paper/paper.tex
Added bibliography and words words words

%\documentclass[aps,prl,preprint,groupedaddress]{revtex4-1}
%\documentclass[aps,prl,preprint,superscriptaddress]{revtex4-1}
\documentclass[aps,prl,reprint,superscriptaddress]{revtex4-1}
\newcommand{\SSC}{S/S_{c}}
\newcommand{\mnras}{Monthly Notices of the Royal Astronomical Society}
\newcommand{\apjs}{The Astrophysical Journal Supplement Series}
\usepackage{graphicx}
% You should use BibTeX and apsrev.bst for references
% Choosing a journal automatically selects the correct APS
% BibTeX style file (bst file), so only uncomment the line
% below if necessary.
%\bibliographystyle{apsrev4-1}

@@

The magnetorotational instability (MRI) is an extremely important instability in astrophysical fluid dynamics.
By changing the stabiliity criterion for differentially rotating flows from a negative angular \emph{momentum} gradient to a negative angular \emph{velocity} gradient, a weak magnetic field allows the Keplerian velocity profile due to a central point mass to drive turbulence \citep[e.g.][]{1998RvMP...70....1B}.
This discovery provided a pathway to explain the ubiquitous accretion onto compact objects at rates compatible with observations and may play a significant role in the formation of planets \citep[e.g.][]{2007Natur.448.1022J}.
In these objects, the gravitational field dominates the local plasma dynamics, and thus the MRI cannot significantly affect the background shear.
It must saturate by other means in these situations \citep{2018MNRAS.474.3451X}.
However, stars have differential rotation profiles driven by much weaker stresses.
In these regions, the MRI could be active and because the driving is much weaker, it can sature by pushing the background shear close to critical \citep{2017ApJ...841....1C,2017ApJ...841....2C}.

Here, we investigate the linear stability of three dimensional perturbations near the critical shear rate $S_c$ for a nearly invisicd, ideal MHD flow.
We find that the first modes to become destabilized are three-dimensional and thus act as a dynamo even in the absence of secondary instability.
These results strongly suggest that the non-normality of the linear operator leading to the MRI are always important, even near onset.
% \begin{itemize}
% \item 3D $\to$ dynamo
% \item stars, NSSL
% \item experiments
% \item Squire's theorem violation? Check MHD Squire, check S instead of Re...
% \item Do eigenvalue solution for B growth? i.e.\ check if it is a dynamo?
% \end{itemize}

We solve the linearized magnetohydrodynamic equations in a frame rotating with angular frequency $\Omega$. The Navier-Stokes equation is in a fairly standard form,
\begin{equation}
  \label{eq:ns}
  \frac{D \mathbf{v}}{Dt} - f \hat{z} \times \mathbf{v} + S v_x \hat{z} + \mathbf{\nabla}{p} + \nu \mathbf{\nabla} \times \mathbf{\omega} = 0,
\end{equation}
where $\omega = \mathbf{\nabla} \times \mathbf{v}$ is the vorticity, $f = 2 \Omega$ is the Coriolis parameter, and $S$ is the background shear rate.
It is common in the accretion disk literature to relate $S$ and $f$ via the $q = S/f$ parameter; for all runs here, $q = 0.75$.
$Df/Dt = \partial f/\partial t + S x \partial f/\partial y$ is the advective derivative including the background linear shear velocity.  We write the induction equation in terms of the $x$ component of the magnetic field,
\begin{equation}
  \label{eq:Bx}
  \frac{D B_x}{Dt} - B_0 \partial_z v_x + \eta (\partial_y J_z - \partial_z J_y) = 0
\end{equation}
and the $x$ component of the current density,
\begin{equation}
  \label{eq:Jx}
  \frac{D J_x}{Dt} - B_0 \partial_z \omega_x + S \partial_z B_x - \eta \nabla^2 J_x = 0
\end{equation}
We explicitly enforce the divergence constraint on the incompressible velocity field $v$ and the magnetic field,
\begin{equation}
  \label{eq:divu}
  \mathbf{\nabla} \cdot \mathbf{v} = \mathbf{\nabla} \cdot \mathbf{B} = 0.
\end{equation}
The magnetic field is expressed in Alfven units, $\mathbf{B} = \mathbf{B_*}/\sqrt{\mu_0 \rho}$.
We consider the behavior of the MRI in a vertically ($z$) and horizontally ($y$) periodic channel of width $d = \pi$. The walls at $x = \pm d/2$ are perfectly conducting and stress-free. The MRI is a weak-field instability; in the inviscid, ideal case the critical shear rate for instability in two dimensions is given by
\begin{equation}
  \label{eq:Sc}
  S_c = \frac{-\pi^2 B^2}{f d^2}.
\end{equation}
Here, we use $\SSC$ as our control parameter.

We solve equations~(\ref{eq:ns})--~(\ref{eq:divu}) by assuming harmonic perturbations in $y$ and $z$, $f = \hat{f}(x) e^{i(k_y y + k_z z) + \sigma t}$, where $\sigma = \gamma + i\omega$ with $\gamma$ and $\omega$ real. This reduces to an eigenvalue problem in $x$, which we solve using the \emph{Dedalus} framework. The advantage of this particular form is that it can be expressed as a set of ten first-order equations with Dirichlet boundary conditions for the stress-free, perfect conductor boundary conditions we use. 
Our interest is in nearly ideal ($\eta = 0$), inviscid ($\nu = 0$) conditions.
However, in order to avoid critical layers in the stable solutions, we set $\eta=\nu=10^{-5}$.
We have confirmed our main results are insensitive to this particular value.

Our main result is that at small values of supercriticality, the fastest growing mode is not axisymmetric. 
Figure~\ref{fig:growth_rate}a shows the growth rates for four values of $\SSC$. 
At $\SSC = 1.02$, the maximum growth rate is found at $(k_y, k_z) \simeq (0.263, 0.447)$.
In fact, at $\SSC = 0.640$, there are no unstable two-dimensional modes, but three-dimensional instability is present.

While surprising, this result can be predicted analytically.
By expanding equations~(\ref{eq:ns}-\ref{eq:divu}) in terms of $S = S_c + \epsilon R$, we compute the leading order correction to the growth rate from three-dimensional effects as
\begin{widetext}
\begin{equation}
  \label{eq:asymp}
\sigma^{2} \ = -\frac{q\,B^{2}}{(q+1)} \left[ k_z^2 \left(\frac{d^2}{\pi^{2}} k_z^2-R\right) \ - \ \frac{\left(6+\pi ^2\right) q+\pi
   ^2-6}{12 } \, k_{y}^{2}  \right] \ + \ ...
\end{equation}
\end{widetext}
The first term in equation~(\ref{eq:asymp}) is the result from a two-dimensional calculation.
The second term is negative definite and proportional to $k_y^2$: for sufficiently large $k_y$, it leads to instability even when the two-dimensional modes are stable.

@@

\begin{figure*}[b!]
  \includegraphics[width=\textwidth]{fig_1.pdf}
  \caption{Stuff}
  \label{fig:growth_rate}
\end{figure*}
\twocolumngrid
%
Numerical simulations have consistently shown axisymmetric modes dominating early evolution of the MRI before parasitic modes cause a breakdown into MHD turbulence \citep{1995ApJ...440..742H,2018ApJ...853..174H,2019ApJS..241...26D}. 
In order to reconcile our results with these simulations, in figure~\ref{fig:alpha} we plot the angle $\alpha = \arctan k_y/k_z$ of the fastest growing mode as function of $\SSC$.
Above $\SSC \gtrsim 2.05$, $\alpha$ is zero, indicating that axisymmetric modes have the fastest growth rates.
For the fiducial run in \citet{1996ApJ...464..690H}, $\SSC \simeq 4.84$ (similar values are found in most works); thus we recover these results.
%
\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{alpha_vs_ssc_grid.pdf}
  \caption{Angle $\alpha$ vs $\SSC$. Stability limit is at $\SSC \simeq 0.102$}
  \label{fig:alpha}
\end{figure}
%
However, even for $\SSC> 2.05$ there are significant swaths of unstable three-dimensional modes with growth rates comparable to the two-dimensional mode (figure~\ref{fig:growth_rate}a).

The linear operator for the MRI, like other shearing systems, is non-normal.
As such, significant non-modal transient amplification occurs even for stable systems.

>>>>

% \documentclass[aps,prl,preprint,groupedaddress]{revtex4-1}
%\documentclass[aps,prl,preprint,superscriptaddress]{revtex4-1}
\documentclass[aps,prl,reprint,superscriptaddress]{revtex4-1}
\newcommand{\SSC}{S/S_{c}}
\newcommand{\mnras}{Monthly Notices of the Royal Astronomical Society}
\newcommand{\apjs}{The Astrophysical Journal Supplement Series}
\usepackage{graphicx}
% You should use BibTeX and apsrev.bst for references
% Choosing a journal automatically selects the correct APS
% BibTeX style file (bst file), so only uncomment the line
% below if necessary.
%\bibliographystyle{apsrev4-1}

@@

The magnetorotational instability (MRI) is an extremely important instability in astrophysical fluid dynamics.
By changing the stabiliity criterion for differentially rotating flows from a negative angular \emph{momentum} gradient to a negative angular \emph{velocity} gradient, a weak magnetic field allows the Keplerian velocity profile due to a central point mass to drive turbulence \citep[e.g.][]{1998RvMP...70....1B}.
This discovery provided a pathway to explain the ubiquitous accretion onto compact objects at rates compatible with observations and may play a significant role in the formation of planets \citep[e.g.][]{2007Natur.448.1022J}.
In these objects, the gravitational field dominates the local plasma dynamics, and thus the MRI cannot significantly affect the background shear.
It must saturate by other means in these situations \citep{2018MNRAS.474.3451X}.
However, stars and liquid metal Taylor-Couette expereiments have differential rotation profiles driven by much weaker stresses.
Where the MRI is active in these flows, it can sature by pushing the background shear close to critical \citep{2017ApJ...841....1C,2017ApJ...841....2C}, analogous to the saturation of convection by relaxing to a nearly constant entropy profile.
Stellar interiors are at extremely high fluid and magnetic Reynolds numbers, but may potentially operate at or near the critical shear rate for the MRI\@.
This critical shear is set by the background magnetic field strength for axisymmetric perturbations in a channel of a given size.
Despite the extensive literature on the accretion disk case of strong shear and high Reynolds numbers and the liquid metal literature on weak shear and lower Reynolds numbers, even the linear dynamics of the MRI are not well studied in the weak shear, high Reynolds number regime.

Here, we investigate the linear stability of three dimensional perturbations near the two dimensional critical shear rate $S_c$ for a nearly invisicd, ideal MHD flow.
We find that the first modes to become destabilized are three-dimensional and thus act as a dynamo even in the absence of secondary instability.
These results strongly suggest that the non-normality of the linear operator driving the MRI are always important, even when axisymmetric modes dominate.
% \begin{itemize}
% \item 3D $\to$ dynamo
% \item stars, NSSL
% \item experiments
% \item Squire's theorem violation? Check MHD Squire, check S instead of Re...
% \item Do eigenvalue solution for B growth? i.e.\ check if it is a dynamo?
% \end{itemize}

We solve the linearized magnetohydrodynamic equations in rotating plane Couette geometry.
This corresponds to a Cartesian frame rotating with angular frequency $\Omega$ and a linear background shear, $v_{y0} = Sx$. 
We cast the Navier-Stokes equation in a fairly standard form,
\begin{equation}
  \label{eq:ns}
  \frac{D \mathbf{v}}{Dt} - f \hat{z} \times \mathbf{v} + S v_x \hat{z} + \mathbf{\nabla}{p} + \nu \mathbf{\nabla} \times \mathbf{\omega} = 0,
\end{equation}
where $\omega = \mathbf{\nabla} \times \mathbf{v}$ is the vorticity, $f = 2 \Omega$ is the Coriolis parameter, and $S$ is the background shear rate.
It is common in the accretion disk literature to relate $S$ and $f$ via the $q = S/f$ parameter; for all runs here, $q = 0.75$.
$Df/Dt = \partial f/\partial t + S x \partial f/\partial y$ is the advective derivative including the background linear shear velocity.  We write the induction equation in terms of the $x$ component of the magnetic field,
\begin{equation}
  \label{eq:Bx}
  \frac{D B_x}{Dt} - B_0 \partial_z v_x + \eta (\partial_y J_z - \partial_z J_y) = 0
\end{equation}
and the $x$ component of the current density,
\begin{equation}
  \label{eq:Jx}
  \frac{D J_x}{Dt} - B_0 \partial_z \omega_x + S \partial_z B_x - \eta \nabla^2 J_x = 0
\end{equation}
We explicitly enforce the divergence constraint on the incompressible velocity field $v$ and the magnetic field,
\begin{equation}
  \label{eq:divu}
  \mathbf{\nabla} \cdot \mathbf{v} = \mathbf{\nabla} \cdot \mathbf{B} = 0.
\end{equation}
The magnetic field is expressed in Alfven units, $\mathbf{B} = \mathbf{B_*}/\sqrt{\mu_0 \rho}$.
We consider the behavior of the MRI in a vertically ($z$) and horizontally ($y$) periodic channel of width $d = \pi$. The walls at $x = \pm d/2$ are perfectly conducting and stress-free. The MRI is a weak-field instability; in the inviscid, ideal case the critical shear rate for instability in two dimensions is given by
\begin{equation}
  \label{eq:Sc}
  S_c = \frac{-\pi^2 B^2}{f d^2}.
\end{equation}
Here, we use $\SSC$ as our control parameter.

We solve equations~(\ref{eq:ns})--~(\ref{eq:divu}) by assuming harmonic perturbations in $y$ and $z$, $f = \hat{f}(x) e^{i(k_y y + k_z z) + \sigma t}$, where $\sigma = \gamma + i\omega$ with $\gamma$ and $\omega$ real. This reduces to an eigenvalue problem in $x$, which we solve using the \emph{Dedalus} framework. The advantage of this particular form is that it can be expressed as a set of ten first-order equations with Dirichlet boundary conditions for the stress-free, perfect conductor boundary conditions we use. 
Our interest is in ideal ($\eta = 0$), inviscid ($\nu = 0$) conditions.
However, in order to avoid critical layers in the stable solutions, we set $\eta=\nu=10^{-5}$.
We have confirmed our main results are insensitive to this particular value. 
1For each $(k_y, k_z)$ pair, we solve an eigenvalue problem with $n_x = 128$ modes; our results are identical at double the resolution\footnote{All code used in this paper can be found at \protect\url{https://github.com/jsoishi/mri\_prefers\_3d}}.

Our main result is that at small values of supercriticality, the fastest growing mode is not axisymmetric. 
Figure~\ref{fig:growth_rate}a shows the growth rates for four values of $\SSC$. 
At $\SSC = 1.02$, the maximum growth rate is found at $(k_y, k_z) \simeq (0.263, 0.447)$.
In fact, the first panel of figure~\ref{fig:growth_rate}a shows $\SSC = 0.640$, where there are no unstable two-dimensional modes, but three-dimensional instability is present.
The critical shear for three dimensional perturbations is at $S_{c,3D} \simeq 0.
102 S_c$.

While surprising, that three-dimensional perturbations grow faster than two-dimensional ones can be predicted analytically.
By expanding equations~(\ref{eq:ns}-\ref{eq:divu}) in terms of $S = S_c + \epsilon R$, such that the shear is asymptotically close to the critical rate for two dimensional perturbations, we compute the leading order correction to the growth rate from three-dimensional effects as
\begin{widetext}
\begin{equation}
  \label{eq:asymp}
\sigma^{2} \ = -\frac{q\,B^{2}}{(q+1)} \left[ k_z^2 \left(\frac{d^2}{\pi^{2}} k_z^2-R\right) \ - \ \frac{\left(6+\pi ^2\right) q+\pi
   ^2-6}{12 } \, k_{y}^{2}  \right] \ + \ ...
\end{equation}
\end{widetext}
The first term in equation~(\ref{eq:asymp}) is the result from a two-dimensional calculation.
The second term is negative definite and proportional to $k_y^2$: for sufficiently large $k_y$, it leads to instability even when the two-dimensional modes are stable.

@@

\begin{figure*}[h!]
  \includegraphics[width=\textwidth]{fig_1.pdf}
  \caption{Stuff}
  \label{fig:growth_rate}
\end{figure*}
\twocolumngrid
%
Numerical simulations have consistently shown axisymmetric modes dominating early evolution of the MRI before breaking down into MHD turbulence \citep{1995ApJ...440..742H,2018ApJ...853..174H,2019ApJS..241...26D}. 
These axisymmetric modes, called ``channel modes,'' are exact non-linear solutions of the governing equations with shearing-periodic boundary conditions in $x$, and thus must saturate via parasitic modes from physics beyond non-linear mode coupling \citep{1994ApJ...432..213G}.
Though we use impenetrable, stress-free boundary conditions that do not allow the unbounded growth of channel modes, the basic principle that axisymmetric disturbences grow fastest is not clearly caused by boundary conditions.
thus, we must reconcile our results with those earlier simulations.
In figure~\ref{fig:alpha}, we plot the angle $\alpha = \arctan k_y/k_z$ of the fastest growing mode with the $z$ axis as function of $\SSC$.
Above $\SSC \gtrsim 2.05$, $\alpha$ is zero, indicating that axisymmetric modes have the fastest growth rates.
For the fiducial run in \citet{1996ApJ...464..690H}, $\SSC \simeq 4.84$ (similar values are found in most works).
Thus, our theory predicts that the linear dynamics should be dominated by axisymmetric modes for the parameters studied in prior numerical simulations.
%
\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{alpha_vs_ssc_grid.pdf}
  \caption{Angle $\alpha$ vs $\SSC$. Stability limit is at $\SSC \simeq 0.102$}
  \label{fig:alpha}
\end{figure}
%

However, even for $\SSC> 2.05$ there are significant swaths of unstable three-dimensional modes with growth rates comparable to the two-dimensional mode (figure~\ref{fig:growth_rate}a).
There are two significant implications for this.
First, the linear operator for the MRI, like other shearing systems, is non-normal; this effect appears whenever non-axisymmetric perturbations are allowed \citep[see][for a discussion relevant to the MRI]{1992MNRAS.255P..25K}.
Because of this, non-modal transient amplification occurs even for stable modes.
These effects can be strong enough to cause non-linear interactions and turbulence even when the entire flow is linearly stable.
While we are here concerned with systems linearly unstable to the MRI, the presence of three dimensional modes implies that linear, non-normal behavior could play a very strong role in the MRI near onset.

The second implication of the three dimensional modes is of considerably more importance.
These modes produce dynamo action in the linear regime, as evidenced by the presence of non-trivial eigenmodes for all three components of the magnetic field.
This is of considerable significance, as it leads to the possibility of a laminar MRI dynamo operating directly in regions of weak shear.
MRI dynamos have been studied in a number of different contexts \citep{2007PhRvL..98y4502R,2011ApJ...740...18O,2015PhRvL.114h5002S}, but all except one have focused on the turbulent dynamo. The lone exception, \citet{2016MNRAS.462..818B}, is a numerical study that discovered the exponential growth of mean magnetic fields during the linear growth phase of the MRI far from stability.
Our work explains this result in terms of purely linear dynamics:
non-axisymmetric MRI unstable modes drive the exponential growth of magnetic fields.
